Effects of the body force on the pedestrian and the evacuation dynamics

### 2020-02-14 paper.tex
tache texto que no va
correcciones de guille sobre las correcciones de dorso

The Social Force Model (SFM) is a suitable model for describing crowd behaviors
\textcolor{blue}{under emotional stress}.
This research analyzes the role of the body force in the original SFM.
We focused on the parameter associated with the body stiffness ($k_n$) and its impact on the pedestrian
dynamics for two different geometries: bottlenecks and corridors.
Increasing $k_n$ produces opposite effects on \textcolor{blue}{crowd dynamics for each geometry}: an increase of the crowd velocity 
for bottlenecks, and a decrease for corridors. 
The former reflects the fact that \textcolor{blue}{an increase in the stiffness reduces
the overlap between pedestrians and, as a consequence, the sliding friction is reduced.} 
This phenomenon reduces the number of blocking clusters close to the exit door. 
\textcolor{blue}{In the case of the corridor, instead, due to the confining walls,
the pedestrians get tight }
 into a lattice-like configuration due to space limitations.
The friction interaction with the walls determines the velocity of the whole crowd along the corridor.
Additionally, the corridor geometry \textcolor{blue}{generates} a flux slowing down for very
crowded environments, as observed in many real-life situations.
We also explored the dimensionless parameters that arose from the reduced-in-units
equation of motion and tuned them to reproduce the qualitative behavior of the empirical fundamental diagram.

@@

The numerical \textcolor{blue}{parameters} for the ``faster is slower'' effect presented by 
Helbing and co-workers \textcolor{red}{\sout{is somewhat cumbersome} has some drawbacks} 
Ref.~\cite{helbing_2000,dorso_2017,dorso_2019}. Although a single 
parameter (say, the desired velocity $v_d$) is numerically varied to 
explain the phenomenon, the researcher \textcolor{red}{\sout{looses sight} may loose sight} of the dimensionless 
``numbers'' that truly control the crowd dynamics. The \textcolor{blue}{set of parameters} in 
Ref.~\cite{helbing_2000} also misses the ``faster is faster'' effect reported 
to occur at very high pedestrian densities \cite{dorso_2017,haghani_2019}. 
\textcolor{blue}{Moreover}, the empirical fundamental diagram raises as a point of reference 
for the SFM control parameters \cite{helbing_2007,dorso_2017}. \\

@@

The complex behavior of pedestrians features
 either his (her) feelings and 
the environmental conditions. The former is expressed, for example, by his 
(her) moving ``attitude'' (say,\textcolor{red}{\sout{self-assuredness} relaxed, normal, aggressive, etc}).
The latter brings out the 
observed separation between pedestrians. Also, the ``contacts'' between 
individuals \textcolor{blue}{produce} some kind of ``unwanted'' slowing down. All these 
observed patterns are commonly quantified in the literature into a set of 
characteristic  parameters. \textcolor{red}{It is worth mentioning that the empirical
data reported in this Section corresponds to non-panicking situations in which pedestrians 
do not experience anxiety. }
The experimental meaning of these parameters is as follows.

\begin{itemize}
\item[(i)] The walking attitude of a pedestrian may appear somewhat 
  ``aggressive'' if he (she) reacts actively to unexpected behaviors 
\cite{lakoba_2005,helbing_1995}. The smaller the reaction time, the more 
 aggressive observed posture. The associated parameter to this 
behavior is the relaxation or characteristic time $\tau$ 
\cite{johansson_2009,helbing_2000}.

\item[(ii)] Despite the reactive attitude $\tau$, the pedestrian \textcolor{blue}{adopts} a 
``free'' (undisturbed) walking speed $v_d$. This speed expresses his (her) 
motivation or intention to reach a certain destination (as comfortable as 
possible). Observations commonly associate $0.6\,$m/s, $1\,$m/s or $1.5\,$m/s 
to relaxed, normal or nervous walking speeds, respectively  
\cite{helbing_1995,helbing_2000,li_2015}. 
\textcolor{red}{Nevertheless, when pedestrians are in a panic situation,
 the actual speed tends to be lower than the desired velocity $v_d$. }

@@

The reports from Refs.~\cite{seyfried_2007,weidmann_1992} do not include any 
values for the compressibility $k_n$ since these experiments were carried out 
under low density conditions.  The minimum (perceived) step distance 
is $0.36\,$m according to Ref.~\cite{seyfried_2007}, but the
 pedestrians seem to require larger 
distances when they walk faster. The commonly accepted value $B\simeq 
0.5\,$m is somewhat valid for walking speeds under $0.5\,$m/s 
\cite{seyfried_2007}. Higher walking speeds (say, $1\,$m/s) will require 
a step distance of $1.3\,$m for the pedestrians to feel that there is enough 
space to move along.\\

@@

The maximum (realistic) compressions may be computed from the \textcolor{blue}{Hook's} relation 
$F(x)=k_n(x)\,x$ and the compressibility $k_n(x)$ reported in Table~\ref{table_data}. 
An ``uncomfortable'' body force  $10\,$N $-$ $30\,$N can address compressions in the 
range of $0.030-0.055\,$m. Also, a ``hitting'' force of $60\,$N can address 
compressions between $0.045$ and $0.065\,$m. \textcolor{red}{
Besides, no reliable values for the sliding friction $k_t$ appears to be 
available in the literature (to our knowledge).}\\
%Thus, according to Table~\ref{table_data}, \textcolor{blue}{we may expect experimental values for $k_n(x)$ up to 
%$1000\,$kg/s$^2$ (for $F=30\,$N), or, up to $1400\,$kg/s$^2$ (for $F=60\,$N).} \\

@@

\noindent where the subscript $j$ corrresponds to any neighboring pedestrian or 
the walls. The three forces $\mathbf{f}_d$, $\mathbf{f}_s$ and $\mathbf{f}_g$ 
are different in nature. The desire force $\mathbf{f}_d$ represents the 
acceleration (or deceleration) of the pedestrian due to his (her) own \textcolor{blue}{will}. 
The social force $\mathbf{f}_s$, instead, describes the tendency of the 
pedestrians to stay away from each other. The granular force $\mathbf{f}_g$ 
stands for either the sliding friction and the compression between 
pedestrians. \\

Notice that these forces are supposed to influence the behavior of the 
pedestrians in a similar fashion as mentioned in Section~\ref{experimental}. 
Thus, the set of (empirical) parameters described in Section~\ref{experimental} 
is expected to be also present in the SFM. These will appear in connection to 
the forces, although their meaning may be somewhat different. \\

@@

The net distance (overlap) $|R_{ij}-r_{ij}|$ scales to the parameter $B$ in the 
expression (\ref{eqn_social}). This parameter plays the role of a fall-off 
length within the model, and thus, it may be somewhat connected to the 
(perceived) step distance mentioned in Section~\ref{experimental}. The 
parameter $A$, however, does not provide any direct link to other parameters 
mentioned there. \\

@@

The numerical \textcolor{blue}{\sout{setting} set} of the parameters may affect the dynamics of the 
pedestrians. Some \textcolor{blue}{\sout{settings} sets}, however, yield similar collective dynamics.
We introduce \textcolor{blue}{dimensionless} magnitudes 
and proceed straightforward as indicated in \ref{appendix1}. We realize from 
\ref{appendix1} that only three (\textcolor{blue}{dimensionless}) parameters  are the true 
``control'' parameters for the collective dynamics. These are $\mathcal{A}$, 
$\mathcal{K}$, $\mathcal{K}_c$ as defined in \ref{appendix1}. Recall that 
$\mathcal{A}$ and $\mathcal{K}$ are precisely the same as in 
Ref.~\cite{dorso_2019}, but a novel $\mathcal{K}_c$ parameter has been 
introduced due to the body force.   \\

@@

Fig.~\ref{venn_2parameters} represents the situation when $\mathcal{K}_c$ is 
absent. Notice that $A/v_d$ or $k_t B$ may control the collective 
dynamic, in spite of the ``assertive'' attitude. The individual character of 
$v_d$ or $B$ appears somehow ``loosely'' in the crowd dynamic. We mean by 
``loosely'' that any numerical \textcolor{blue}{\sout{setting} set} of these parameters may be 
counterbalanced by the right \textcolor{blue}{\sout{setting} choise } of $A$ or $k_t$, keeping the 
colletive dynamic (qualitatively) unchanged.  \\

@@

\item[(b)] We will consider the parameter \textcolor{blue}{set of} Ref.~\cite{helbing_2000} 
as a starting point for exploring the parameter space. The reason for this is 
that a complete set of experimental parameters is still not available (to our 
konwledge). We are aware, though, of the drawbacks \textcolor{blue}{of} this choise, say, the 
unrealistic meanings for the length $B$ and the compressibility $k_n$ (see 
Section~\ref{experimental}). But these are irrelevant in the context of the 
(\textcolor{blue}{dimensionless}) parameter space and the corresponding collective dynamics (see 
Section~\ref{introduction}).

@@

In order to keep the model as simple as possible, we will only consider 
\textcolor{blue}{isotropic pedestrian interactions}, since \textcolor{blue}{anisotropic
interactions} does not play a relevant role at high densities. \\

@@

The implemented SFM parameters were the same as those in 
Ref.~\cite{helbing_2000} (at the beginning of the exploratory procedure only). 
But the pedestrian's mass and radius were set to the more realistic values of 
$70\,$kg and $0.23\,$m, respectively. The force interactions between 
pedestrians were limited, however, to a cut-off distance of $0.88\,$m for 
attaining a \textit{privacy sphere} \textcolor{blue}{only of the first neighbors}. The desired 
velcity was always set to $1\,$m/s in the corridor situation. Besides, the 
explored values of $v_d$ for the bottleneck scenario ranged from $1\,$m/s to the 
extremely anxious situation of $10\,$m/s. \\

@@

Notice that the mean degree increases as $v_d$ increases, as expected. This expresses the fact that
higher $v_d$ values accomplish higher densities, \textcolor{blue}{forcing} individuals to touch each 
other. For a given $v_d$, the mean degree reduces as the $k_n$ 
value increases. A noticeable decrease in the mean degree can be seen for the 
highest explored value of $k_n$. This \textcolor{blue}{was not expected and} opens the question on how would
this affect the sliding friction among pedestrians.\\

>>>>

The Social Force Model (SFM) is a suitable model for describing crowd behaviors
\textcolor{blue}{under emotional stress}.
This research analyzes the role of the body force in the original SFM.
We focused on the parameter associated with the body stiffness ($k_n$) and its impact on the pedestrian
dynamics for two different geometries: bottlenecks and corridors.
Increasing $k_n$ produces opposite effects on \textcolor{blue}{the crowd dynamics for each geometry}: an increase of the crowd velocity 
for bottlenecks, and a decrease for corridors. 
The former reflects the fact that \textcolor{blue}{an increase in the stiffness reduces
the overlap between pedestrians and, as a consequence, the sliding friction is reduced.} 
This phenomenon reduces the number of blocking clusters close to the exit door. 
\textcolor{blue}{In the case of the corridor, instead, due to the confining walls,
the pedestrians get tight }
 into a lattice-like configuration due to space limitations.
The friction interaction with the walls determines the velocity of the whole crowd along the corridor.
Additionally, the corridor geometry \textcolor{blue}{generates} a flux slowing down for very
crowded environments, as observed in many real-life situations.
We also explored the dimensionless parameters that arose from the reduced-in-units
equation of motion and tuned them to reproduce the qualitative behavior of the empirical fundamental diagram.

@@

The numerical \textcolor{blue}{parameters} for the ``faster is slower'' effect presented by 
Helbing and co-workers \textcolor{red}{\sout{is somewhat cumbersome} appears to have some drawbacks} 
Ref.~\cite{helbing_2000,dorso_2017,dorso_2019}. Although a single 
parameter (say, the desired velocity $v_d$) is numerically varied to 
explain the phenomenon, the researcher \textcolor{red}{\sout{looses sight} may loose sight} of the dimensionless 
``numbers'' that truly control the crowd dynamics. The \textcolor{blue}{set of parameters} in 
Ref.~\cite{helbing_2000} also misses the ``faster is faster'' effect reported 
to occur at very high pedestrian densities \cite{dorso_2017,haghani_2019}. 
\textcolor{blue}{Moreover}, the empirical fundamental diagram raises as a point of reference 
for the SFM control parameters \cite{helbing_2007,dorso_2017}. \\

@@

The complex behavior of pedestrians features
 either his (her) feelings and 
the environmental conditions. The former is expressed, for example, by his 
(her) moving ``attitude'' (say,\textcolor{red}{\sout{self-assuredness} relaxed, normal, aggressive, etc}).
The latter brings out the 
observed separation between pedestrians. Also, the ``contacts'' between 
individuals \textcolor{blue}{produce} some kind of ``unwanted'' slowing down. All these 
observed patterns are commonly quantified in the literature into a set of 
characteristic  parameters. \textcolor{red}{It is worth mentioning that the empirical
data reported in this Section corresponds to non-panicking situations in which pedestrians 
do not experience high anxiety levels (see below). }
The experimental meaning of these parameters is as follows.

\begin{itemize}
\item[(i)] \textcolor{red}{\textbf{The walking attitude of a pedestrian may appear somewhat 
  ``aggressive'' if he (she) reacts actively to unexpected behaviors 
\cite{lakoba_2005,helbing_1995}. The smaller the reaction time, the more 
 aggressive observed posture. The associated parameter to this 
behavior is the relaxation or characteristic time $\tau$ 
\cite{johansson_2009,helbing_2000}}}.

\item[(ii)] Despite the reactive attitude $\tau$, the pedestrian \textcolor{blue}{adopts} a 
``free'' (undisturbed) walking speed $v_d$. This speed expresses his (her) 
motivation or intention to reach a certain destination (as comfortable as 
possible). Observations commonly associate $0.6\,$m/s, $1\,$m/s or $1.5\,$m/s 
to relaxed, normal or nervous walking speeds, respectively  
\cite{helbing_1995,helbing_2000,li_2015}. 
\textcolor{red}{Nevertheless, when pedestrians are in a panic situation,
 they may not reach the ``free'' walking speed, but, $v_d$ may still be 
 associated to his (her) motivation to get to a target position.  }

@@

The reports from Refs.~\cite{seyfried_2007,weidmann_1992} do not include any 
values for the compressibility $k_n$ since these experiments were carried out 
under low density \textcolor{red}{and non-panicking} conditions.  The minimum (perceived) step distance 
is $0.36\,$m according to Ref.~\cite{seyfried_2007}, but the
 pedestrians seem to require larger 
distances when they walk faster. The commonly accepted value $B\simeq 
0.5\,$m is somewhat valid for walking speeds under $0.5\,$m/s  \textcolor{red}{and non-panicking conditions}
\cite{seyfried_2007}. Higher walking speeds (say, $1\,$m/s) will require 
a step distance of $1.3\,$m for the pedestrians to feel that there is enough 
space to move along.\\

@@

The maximum (realistic) compressions may be computed from the \textcolor{blue}{Hooke's} relation 
$F(x)=k_n(x)\,x$ and the compressibility $k_n(x)$ reported in Table~\ref{table_data}. 
An ``uncomfortable'' body force  $10\,$N $-$ $30\,$N can address compressions in the 
range of $0.030-0.055\,$m. Also, a ``hitting'' force of $60\,$N can address 
compressions between $0.045$ and $0.065\,$m. \textcolor{red}{
Besides, no reliable values for the sliding friction $k_t$ appears to be 
available in the literature (to our knowledge).}\\
%Thus, according to Table~\ref{table_data}, \textcolor{blue}{we may expect experimental values for $k_n(x)$ up to 
%$1000\,$kg/s$^2$ (for $F=30\,$N), or, up to $1400\,$kg/s$^2$ (for $F=60\,$N).} \\

@@

\noindent where the subscript $j$ corrresponds to any neighboring pedestrian or 
the walls. The three forces $\mathbf{f}_d$, $\mathbf{f}_s$ and $\mathbf{f}_g$ 
are different in nature. The desire force $\mathbf{f}_d$ represents the 
acceleration (or deceleration) of the pedestrian due to his (her) own \textcolor{blue}{will}. 
The social force $\mathbf{f}_s$, instead, describes the tendency of the 
pedestrians to stay away from each other. The granular force $\mathbf{f}_g$ 
stands for \textcolor{red}{both} the sliding friction and the compression between 
pedestrians. \\

Notice that these forces are supposed to influence the behavior of the 
pedestrians in a similar fashion as mentioned in Section~\ref{experimental}. 
Thus, the set of (empirical) parameters described in Section~\ref{experimental} 
is expected to be also present in the SFM. These will appear in connection to 
the forces \textcolor{red}{\sout{, although their meaning may be somewhat different}}. \\

@@

The net distance (overlap) $|R_{ij}-r_{ij}|$ scales to the parameter $B$ in the 
expression (\ref{eqn_social}). This parameter plays the role of a fall-off 
length within the model, and thus, it may be somewhat connected to the 
(perceived) step distance mentioned in Section~\ref{experimental}. 
\textcolor{red}{Besides, the parameter $A$ reflects the intensity of the social repulsion.} \\

@@

The numerical \textcolor{blue}{\sout{setting} set} of the parameters may affect the dynamics of the 
pedestrians. Some \textcolor{blue}{\sout{settings} values}, however, yield similar collective dynamics.
We introduce \textcolor{blue}{dimensionless} magnitudes 
and proceed straightforward as indicated in \ref{appendix1}. We realize from 
\ref{appendix1} that only three (\textcolor{blue}{dimensionless}) parameters  are the true 
``control'' parameters for the collective dynamics. These are $\mathcal{A}$, 
$\mathcal{K}$, $\mathcal{K}_c$ as defined in \ref{appendix1}. Recall that 
$\mathcal{A}$ and $\mathcal{K}$ are precisely the same as in 
Ref.~\cite{dorso_2019}, but a novel $\mathcal{K}_c$ parameter has been 
introduced due to the body force.   \\

@@

Fig.~\ref{venn_2parameters} represents the situation when $\mathcal{K}_c$ is 
absent. Notice that $A/v_d$ or $k_t B$ may control the collective 
dynamic, in spite of the ``assertive'' attitude. The individual character of 
$v_d$ or $B$ appears somehow ``loosely'' in the crowd dynamic. We mean by 
``loosely'' that any numerical \textcolor{blue}{\sout{setting} set for} these parameters may be 
counterbalanced by the right \textcolor{blue}{\sout{setting} choise } of $A$ or $k_t$, keeping the 
colletive dynamic (qualitatively) unchanged.  \\

@@

\item[(b)] We will consider the parameter \textcolor{blue}{set from} Ref.~\cite{helbing_2000} 
as a starting point for exploring the parameter space. The reason for this is 
that a complete set of experimental parameters is still not available (to our 
konwledge). We are aware, though, of the drawbacks \textcolor{blue}{for} this choise, say, the 
unrealistic meanings for the length $B$ and the compressibility $k_n$ (see 
Section~\ref{experimental}). But these are irrelevant in the context of the 
(\textcolor{blue}{dimensionless}) parameter space and the corresponding collective dynamics (see 
Section~\ref{introduction}).

@@

In order to keep the model as simple as possible, we will only consider 
\textcolor{blue}{isotropic pedestrian interactions}, since \textcolor{blue}{anisotropic
interactions do} not play a relevant role at high densities. \\

@@

The implemented SFM parameters were the same as those in 
Ref.~\cite{helbing_2000} (at the beginning of the exploratory procedure only). 
But the pedestrian's mass and radius were set to the more realistic values of 
$70\,$kg and $0.23\,$m, respectively. The force interactions between 
pedestrians were limited, however, to a cut-off distance of $0.88\,$m for 
attaining a \textit{privacy sphere} \textcolor{blue}{for the first neighbors only}. The desired 
velcity was always set to $1\,$m/s in the corridor situation. Besides, the 
explored values of $v_d$ for the bottleneck scenario ranged from $1\,$m/s to the 
extremely anxious situation of $10\,$m/s. \\

@@

Notice that the mean degree increases as $v_d$ increases, as expected. This expresses the fact that
higher $v_d$ values accomplish higher densities, \textcolor{blue}{forcing} individuals to touch each 
other. For a given $v_d$, the mean degree reduces as the $k_n$ 
value increases. A noticeable decrease in the mean degree can be seen for the 
highest explored value of $k_n$. This \textcolor{blue}{\sout{was not expected and}} opens the question on how would
this affect the sliding friction among pedestrians.\\
